---
title: 'STAT 206 Lab 2'
output: pdf_document
name: Casey Kongpanickul
---

**Due Monday, October 16, 5:00 PM**

***General instructions for labs***: You are encouraged to work in pairs to complete the lab.  Labs must be completed as an R Markdown file.  Be sure to include your lab partner (if you have one) and your own name in the file.  Give the commands to answer each question in its own code block, which will also produce plots that will be automatically embedded in the output file. Each answer must be supported by written statements as well as any code used. 

***Agenda***: Manipulating data frames; practicing iteration; practicing re-writing code; checking how reliable random methods are.

Part I -- Data Frames
======

R includes a number of pre-specified data objects as part of its default installation. We will load and manipulate one of these, a data frame of 93 cars with model year 1993. Begin by ensuring that you can load this data with the commands

```{r}
library(MASS)
data(Cars93)
```

Begin by examining the data frame with the command `View(Cars93)` to understand the underlying object. You will need to use functions and other commands to extract elements for this assignment.

```{r}

# Class displays what type a given argument is

class(Cars93)

#View puts the data frame into its own table with header

View(Cars93) 

```


1. Obtain a `summary()` of the full data structure. Can you tell from this how many rows are in the data? If so, say how; if not, use another method to obtain the number of rows.

```{r}

#the summary function returns some descriptve statistics on from the data frame. It includes the "5 number summary"
#the 5 number summary includes the min value, q1, median, q3 and max value
#From this data we can get an idea of how many rows and columns. We can count the number of attrributes that 
#summary returns and we can know the amount of columns, and for the rows we can find an attribute that can be used as 
#an idex to find the number of rows. in this case we see both Model and Make display 6 values and shows 87 other
#My guess would be that there is 93 rows
#It is probably much better to use a couple funcions to accomplish this and display the number of columns and rows

summary(Cars93)

nrow(Cars93)

ncol(Cars93)

# nrow and ncol will display the number of rows and columns in  a dataframe, respectively. This is a much cleaner way to get these numbers compared to extracting them manually from the summary function

```

2. What is the mean price of a car with a rear-wheel drive train?

```{r}

#First step is to get a slice of this data frame to extract only rows that have 'Rear' for attribute DriveTrain
#We also want to get the price values assocaited with the above values
#we can use the subset function for anintuitive way to obtain this slice

rwd_avg_p = subset(Cars93, DriveTrain == 'Rear', select = c('Price', 'DriveTrain'))

#Finally, we use the mean function but only on the price values in our subset

mean(rwd_avg_p[,'Price'])

```


3. What is the minimum horsepower of all cars with capacity for 7 passengers? With a capacity of at least 6 passengers?

```{r}

#Again the subset function will be used to grab a slice of the data frame that consists of horse power and capacity
#first will be a slice for when passengers = 7

cap_hp = subset(Cars93, Passengers == 7, select = c('Passengers', 'Horsepower'))

#the min function is called on the attribute Horsepower for the subset data

min(cap_hp[,'Horsepower'])

#Next we will get a slice for when there are atleast 6 passengers

cap_hp2 = subset(Cars93, Passengers >= 6, select = c('Passengers', 'Horsepower'))

#now we will use the min function again to get the new minimum of this slice

min(cap_hp2[,'Horsepower'])

```

4. Assuming that these cars are exactly as fuel efficient as this table indicates, find the cars that have the maximum, minimum and median distance travellable for highway driving.  You will need at least two columns to work this out; why those two?

```{r}

# To obtain which cars have the best, worst and middle hwy mpg we need to first know what these values are
#we can do this using the max, min, and median functions on the Cars93 data, on the hwy mpg column
# note this info is available in the summary above

maxmpg = max(Cars93[,'MPG.highway'])
maxmpg

medianmpg = median(Cars93[,'MPG.highway'])
medianmpg

minmpg = min(Cars93[,'MPG.highway'])
minmpg

# Now we can create slices of the data frame for Cars that have the above values

maxhighway = subset(Cars93, MPG.highway == maxmpg, select = c('MPG.highway', 'Make'))
maxhighway

medianhighway = subset(Cars93, MPG.highway == medianmpg, select = c('Make', 'MPG.highway'))
medianhighway

minhighway = subset(Cars93, MPG.highway == minmpg, select = c('Make', 'MPG.highway'))
minhighway

```

Part II -- Reproducibility and Functions
==========================================

Some of the lectures have included examples of planning production for a factory that turns steel and labor into cars and trucks.  Below is a piece of code that optimizes the factory's output (roughly) given the available resources, using a `repeat` loop.  It's embedded in a function to make it easier for you to run.

```{r}
factory.function <- function (cars.output=1, trucks.output=1) {
  factory <- matrix(c(40,1,60,3),nrow=2,
    dimnames=list(c("labor","steel"),c("cars","trucks")))
  available <- c(1600,70); names(available) <- rownames(factory)
  slack <- c(8,1); names(slack) <- rownames(factory)
  output <- c(cars.output, trucks.output); names(output) <- colnames(factory)

  passes <- 0 # How many times have we  been around the loop?
  repeat {
     passes <- passes + 1
     needed <- factory %*% output # What do we need for that output level?
     # If we're not using too much, and are within the slack, we're done
     if (all(needed <= available) &&
         all((available - needed) <= slack)) {
       break()
     }
     # If we're using too much of everything, cut back by 10%
     if (all(needed > available)) {
       output <- output * 0.9
       next()
     }
     # If we're using too little of everything, increase by 10%
     if (all(needed < available)) {
       output <- output * 1.1
       next()
     }
     # If we're using too much of some resources but not others, randomly
     # tweak the plan by up to 10%
      # runif == Random number, UNIFormly distributed, not "run if"
     output <- output * (1+runif(length(output),min=-0.1,max=0.1))
  }

  return(output)
}
```

5. Run the function above with the command

```{r}
factory.function()
```

to obtain a default output value, starting from a very low initial planned output. What is the final output capacity obtained?

6. Repeat this four more times to obtain new output values. Do these answers differ from each other? If so why? If not, why not?

```{r}

factory.function()

factory.function()

factory.function()

factory.function()

#All of the answers are relatively close to one another however, all of them are different. This is due to the
#function itself. It even says that it will "randomly tweak the plan by up to %10" This means that each run of
#the function will return a different output

```

7. Right now, the number of `passes` is a value held within the function itself and not shared. Change the code so that the number of `passes` will be returned at the end of the function, as well as the final `output`. 

```{r}

#The return(output) at the end was commented out so I could use a print statement to display the output follow by the number of passes used

factory.function <- function (cars.output=1, trucks.output=1) {
  factory <- matrix(c(40,1,60,3),nrow=2,
    dimnames=list(c("labor","steel"),c("cars","trucks")))
  available <- c(1600,70); names(available) <- rownames(factory)
  slack <- c(8,1); names(slack) <- rownames(factory)
  output <- c(cars.output, trucks.output); names(output) <- colnames(factory)

  passes <- 0 # How many times have we  been around the loop?
  repeat {
     passes <- passes + 1
     needed <- factory %*% output # What do we need for that output level?
     # If we're not using too much, and are within the slack, we're done
     if (all(needed <= available) &&
         all((available - needed) <= slack)) {
       break()
     }
     # If we're using too much of everything, cut back by 10%
     if (all(needed > available)) {
       output <- output * 0.9
       next()
     }
     # If we're using too little of everything, increase by 10%
     if (all(needed < available)) {
       output <- output * 1.1
       next()
     }
     # If we're using too much of some resources but not others, randomly
     # tweak the plan by up to 10%
      # runif == Random number, UNIFormly distributed, not "run if"
     output <- output * (1+runif(length(output),min=-0.1,max=0.1))
     
     
  }

  #return(output)
  print(output) + print(passes)
}

factory.function()

```

8. Now, set the initial output levels to 30 cars and 20 trucks and run the code.  What is the final output plan (`output`)?  What is the final demand for resources (`needed`)?  Is the plan within budget and within the slack?  How many iterations did it take to converge (`passes`)? For all but `output` you will need to either print this message out deliberately, or return an object that contains all the quantities you want.


```{r}

# the code was copied from aove but in the print statement we are printing out 'needed' to see how much of the
#material was consumed for the output. In this case the output was almost 10 cars and 20 trucks

factory.function <- function (cars.output=1, trucks.output=1) {
  factory <- matrix(c(40,1,60,3),nrow=2,
    dimnames=list(c("labor","steel"),c("cars","trucks")))
  available <- c(1600,70); names(available) <- rownames(factory)
  slack <- c(8,1); names(slack) <- rownames(factory)
  output <- c(cars.output, trucks.output); names(output) <- colnames(factory)

  passes <- 0 # How many times have we  been around the loop?
  repeat {
     passes <- passes + 1
     needed <- factory %*% output # What do we need for that output level?
     # If we're not using too much, and are within the slack, we're done
     if (all(needed <= available) &&
         all((available - needed) <= slack)) {
       break()
     }
     # If we're using too much of everything, cut back by 10%
     if (all(needed > available)) {
       output <- output * 0.9
       next()
     }
     # If we're using too little of everything, increase by 10%
     if (all(needed < available)) {
       output <- output * 1.1
       next()
     }
     # If we're using too much of some resources but not others, randomly
     # tweak the plan by up to 10%
      # runif == Random number, UNIFormly distributed, not "run if"
     output <- output * (1+runif(length(output),min=-0.1,max=0.1))
     
     
  }

  #return(output)
  print(output) + print(passes) + print(needed)
}

factory.function(cars.output = 1, trucks.output = 20)

```
